Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations

### 2021-11-09 analysis/paper/paper.Rmd
added map and updated site descriptions

The Deadwater Slough is located in a reach of the Salmon River that is believed to be a historically important overwinter rearing area for DSR emigrants. Moreover, this reach is part of the migratory pathway for juvenile DSR and NRR emigrants from all eight extant populations. The slough lacks hydrological and structural features (i.e., homogeneous channel with fine substrate) that can provide essential refuge from predation. As a result, predation on juvenile Chinook salmon proximal to the Deadwater Slough has been cited as a concern for the Upper Salmon River MPG, impacting DSR migrants in the fall and NRR emigrants during the spring.

@@

<!-- The following was pulled directly from a technical report, but probably could/should be improved for the manuscript -->

The Deadwater Slough is an approximately two-kilometer (km) section of the mainstem Salmon River locations roughly 5.8 river kms downstream from the town of North Fork, Idaho ***(publication quality map referenced here)***. The downstream end of the Deadwater Slough is located at the confluence of Dump Creek and the Salmon River. Around 1897, the failure of a small mining diversion reservoir in the Dump Creek drainage resulted in an erosion event that deposited substantial amounts of sediment at the confluence of the Salmon River and Dump Creek [@Emerson1973]. The influx of sediment created a delta that impeded flow of the Salmon River, thereby creating an unnaturally slow and deep section in the river, spanning approximately 25.3 acres. The slower water velocities, deepened channel, and (anecdotally) warmer water temperatures favor piscivorous fishes known to occupy the reach (e.g., northern pikeminnow *Ptychocheilus oregonensis*, smallmouth bass *Micropterus dolomieu*).

@@

# References

@@

>>>>

The Deadwater Slough is located in a reach of the Salmon River that is believed to be a historically important overwinter rearing area for DSR emigrants. Moreover, this reach is part of the migratory pathway for juvenile DSR and NRR emigrants from all eight extant populations. The slough lacks hydrological and structural features (i.e., a homogenous channel with fine substrate and little cover) that can provide essential refuge from predation. As a result, predation on juvenile Chinook salmon proximal to the Deadwater Slough has been cited as a concern for the Upper Salmon River MPG, impacting DSR migrants in the fall and NRR emigrants during the spring.

@@

The Deadwater Slough is an approximately 1.5 kilometer section of the mainstem Salmon River located roughly 5.8 river kilometers downstream from the town of North Fork, Idaho ***(publication quality map referenced here)***. The downstream end of the slough is located at the confluence of Dump Creek and the Salmon River. Around 1897, the failure of a small mining diversion reservoir in the Dump Creek drainage resulted in an erosion event that deposited substantial amounts of sediment at the confluence of the Salmon River, thereby creating an unnaturally slow and deep section in the river, spanning approximately 30 acres and averaging 68 m width. Both northern pikeminnow *Ptychocheilus oregonensis* and smallmouth bass *Micropterus dolomieu* inhabit the slough which likely provides favorable conditions for their feeding and growth (e.g., reduced water velocity, deep channel, warmer water temperature).

@@

# Literature Cited

@@

# Tables

<div id="tables"></div>

# Figures

<div id="figures"></div>

```{r map, out.width = "85%", fig.cap = "Map of the Deadwater Slough study area. The high-resolution orthoimage portion directly surrounding the Deadwater Slough was generated from aerial images taken by a drone. The red polygon indicates the reach characterized by unnaturally slow water velocities and a deepened channel. The location of the Dump Creek delta is indicated."}
knitr::include_graphics("../figures/deadwater_map.png")
